Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-13 write-ups/PLDI/paper.tex
Just basic ideas

Suppose a workgroup with id $m$ executes $\offerkill$.  
If the workgroup has the highest largest id of all active workgroups then it can be killed by the scheduler, except that workgroup 0 can never be killed (to ensure that the kernel computation does not terminate early).  More formally, if $m < M-1$
or $M=1$, i.e.\ the workgroup's id is not the largest among the active
workgroups, or if only a single workgroup is executing the cooperative
kernel, then $\offerkill$ is a no-op.  If instead $M > 1$ and $m =
M-1$, there is a non-deterministic choice between (a) $\offerkill$
executing as a no-op, or (b) execution of the workgroup ceasing and
the number of active workgroups $M$ being atomically decremented by one.  Case
(b) corresponds to the scheduler accepting the workgroup's offer to
cease execution, allowing the scheduler to schedule some other task on
the compute unit on which the workgroup was executing.

@@

The $\offerfork$ primitive (also a workgroup-level function) allows the cooperative kernel to indicate
to the scheduler that it could benefit from, and is prepared to
handle, additional workgroups joining the computation.  Recall that
the cooperative kernel was launched with a limit of $N$ workgroups,
indicating the number of workgroups that would ideally execute the
kernel.  However, the number of active workgroups, $M$, may be smaller
than $N$, either because (due to competing workloads) the scheduler
did not provide $N$ workgroups initially, or because the kernel has
given up some number of workgroups via $\offerkill$ calls.  Through
$\offerfork$, the kernel and scheduler can work together to allow new
workgroups to join the computation at an appropriate point and with
appropriate state.

Suppose a workgroup with id $m\leq M$ executes $\offerfork$.  Then the following occurs: an
integer $k$ in the range $[0, N-M]$ is chosen by the scheduler;
$k$ new workgroups are spawned with consecutive ids in the range $[M,
  M+k-1]$; the active workgroup count $M$ is atomically incremented by $k$.

@@

Semantically, when all threads have reached $\resizingglobalbarrier$,
the number of active workgroups, $M$, is atomically set to a new value, $M'$ say, with $M' > 0$.
If $M' = M$ then the active workgroups remain unchanged and proceed to
execute after the barrier.  If $M' < M$, workgroups $[M', M-1]$ are
killed.  If $M' > M$ then $M'-M$ new workgroups join the computation,
as if they were forked from workgroup 0.  In particular, the
$\transmit$-annotated local state of thread 0 in workgroup 0 is
transmitted to the threads of the new workgroups.  The change in
number of active workgroups from $M$ to $M'$ is atomic.

@@

The enclosing $\globalbarrier$ calls ensure that the change in number
of active workgroups from $M$ to $M'$ occurs entirely within the
resizing barrier, so that from a programmer's perspective, $M$ changes atomically.  The middle $\globalbarrier$ is required to ensure that forking occurs before killing, so that while the barrier may kill a workgroup with id $i$, it should not also fork a new workgroup with id $i$.  That is, workgroups $[0, \mathit{min}(M, M') - 1]$ should be left intact.

@@

The programming model and associated language constructs were inspired
by the requirements of the irregular GPU algorithms proposed in the
literature.  For these algorithms, it was usually easy to identify
where to place the constructs.

@@

\paragraph{A changing number of workgroups}  Unlike in regular OpenCL,
the value returned by $\getnumgroups()$ is not fixed during the
lifetime of a cooperative kernel: it may change due to workgroups executing
$\offerkill$, $\offerfork$ and $\resizingglobalbarrier$.  Because $\getglobalsize()$, which retrieves the total number of threads executing the kernel, is
a function of $\getnumgroups()$, the value it returns is also subject to change.
The programmer
must therefore write their cooperative kernel in a manner that is
robust to changes in the values returned by these functions.  In our experience this either involves avoiding the use of $\getnumgroups()$ at all, or writing a kernel where the only opportunity for the number of groups to change comes from $\resizingglobalbarrier()$, in which case the number of active groups does offer some stability, as discussed below.

@@

For the work stealing example, there is no state to transmit since a
computation is entirely parameterised by a task, which is retrieved from
a task pool located in global memory. In the main loop, before trying to
obtain a task, we make the workgroup offer itself to be forked or killed
(lines~\ref{line:cwksteal:offerfork},\ref{line:cwksteal:offerkill}). Note
that a workgroup may be killed even if its associated task pool is not
empty, since remaining tasks could eventually be stolen by other
workgroups.  Care must be taken on computing pool identifiers: the
original version in \myfig\ref{fig:workstealing} can rely on the fact
that there are as many pools as workgroups, but here the number of
workgroups may change during execution. Therefore, the number of pools
must now be explicitely given as an argument
(line~\ref{line:cwksteal:argnumpool}) and used as an upper bound of pool
indexes when scanning pools in $\mathsf{more\_work}$
(line~\ref{line:cwksteal:morework}) and $\mathsf{pop\_or\_steal}$
(line~\ref{line:cwksteal:poporsteal}). Besides, computation of the
workgroup's pool index (line~\ref{line:cwksteal:getgroupid}) must not
only be bounded by the number of pools but also be moved inside the main
loop since a forked workgroup that starts to execute after $\offerfork$
must re-compute its pool index.

@@

The situation is more stable if a cooperative kernel does not directly
issue $\offerkill$ and $\offerfork$ calls directly.  In this case,
only calls to $\resizingglobalbarrier$ can cause the number of active
workgroups to change.  At any point
during execution, the threads of a kernel are executing between some
pair of resizing barrier calls, which we call a \emph{resizing barrier
  interval} (here we consider the kernel entry and exit points
conceptually to be special cases of resizing barriers).  The number of
workgroups executing the kernel is thus constant within each resizing
barrier interval, i.e.\ the result returned by repeated calls to
$\getnumgroups()$ will be identical during such an interval; the same holds for $\getglobalsize()$.

@@

Recall that the cooperative kernel would ideally utilise the $N$
workgroups requested upon kernel launch, in order to conduct its
processing efficiently.  The scheduler should thus aim to provide with
cooperative kernel with as many workgroups, up to this number, as
other constraints allow.  As a result, the scheduler should generally
accept an $\offerkill$ only if a compute unit is required for another
task, and should respond positively to $\offerfork$ calls by
scheduling new workgroups if compute units are available.  Our
implementation of cooperative scheduling (\mysec\ref{sec:implementation}) is as
generous as possible to the cooperative kernel in this regard.

However, these are not hard-and-fast rules: a scheduler might employ
heuristics based on historical workloads to avoid thrashing.  For
example, if graphics calls are known to arrive at a certain rate then
the scheduler might decide that it is worth accepting an $\offerkill$
from the cooperative kernel on the assumption that a graphics task
will be requested momentarily.  Similarly, the scheduler might provide
some, but not all, compute units to the cooperative kernel in response
to an $\offerfork$ call if it has reason to believe that the remaining
compute units are likely to be useful for other high priority tasks
that are likely to arrive imminently.  With some additional language extensions, it would be possible for the cooperative kernel to hint to the scheduler if, for example, the kernel is at a point of computation where there is less work to do than usual, so that $\offerkill$ calls could be accepted in order to save energy.

@@

\ADComment{Better section title?}

@@

\ADComment{Sreepathi mentioned the NCCL framework.}

>>>>

Suppose a workgroup with id $m$ executes $\offerkill$.  If $m < M-1$
or $M=1$, i.e.\ the workgroup's id is not the largest among the active
workgroups, or if only a single workgroup is executing the cooperative
kernel, then $\offerkill$ is a no-op.  If instead $M > 1$ and $m =
M-1$, there is a non-deterministic choice between (a) $\offerkill$
executing as a no-op, or (b) execution of the workgroup ceasing and
the number of active workgroups $M$ being decremented by one.  Case
(b) corresponds to the scheduler accepting the workgroup's offer to
cease execution, allowing the scheduler to schedule some other task on
the compute unit on which the workgroup was executing.

The semantics of $\offerkill$ is atomic.  That is, if multiple
workgroups reach $\offerkill$ simultaneously, the semantics is as if
each workgroup executes $\offerkill$ in some arbitrary sequential
order.

In this formulation of cooperative kernels, only the active workgroup
with the largest id can be killed, and workgroup 0 can never be
killed.  We discuss the rationale for these choices, as well as
possible alternative choices, in
\mysec\ref{sec:semanticalternatives}.

@@

The $\offerfork$ primitive allows the cooperative kernel to indicate
to the scheduler that it could benefit from, and is prepared to
handle, additional workgroups joining the computation.  Recall that
the cooperative kernel was launched with a limit of $N$ workgroups,
indicating the number of workgroups that would ideally execute the
kernel.  However, the number of active workgroups, $M$, may be smaller
than $N$, either because (due to competing workloads) the scheduler
did not provide $N$ workgroups initially, or because the kernel has
given up some number of workgroups via $\offerkill$ calls.  Through
$\offerfork$, the kernel and scheduler can work together to allow new
workgroups to join the computation at an appropriate point and with
appropriate state.

Like $\offerkill$, $\offerfork$ is a workgroup-level function.

Suppose a workgroup with id $m\leq M$ executes $\offerfork$.  An
integer $k$ in the range $[0, N-M]$ is chosen nondeterministically,
$k$ new workgroups are spawned with consecutive ids in the range $[M,
  M+k-1]$, and the active workgroup count $M$ is incremented by $k$.

@@

Semantically, when all threads have reached $\resizingglobalbarrier$,
the number of active workgroups, $M$, is set to a new value, $M'$ say, with $M' > 0$.
If $M' = M$ then the active workgroups remain unchanged and proceed to
execute after the barrier.  If $M' < M$, workgroups $[M', M-1]$ are
killed.  If $M' > M$ then $M'-M$ new workgroups join the computation,
as if they were forked from workgroup 0.  In particular, the
$\transmit$-annotated local state of thread 0 in workgroup 0 is
transmitted to the threads of the new workgroups.  The change in
number of active workgroups from $M$ to $M'$ is atomic.

@@

The enclosing $\globalbarrier$ calls make the operation appear to be
atomic.  The middle $\globalbarrier$ is required to ensure that if
$M'\geq M$, the workgroups with ids smaller than $M$ are left intact.

@@

The programming model and associated language constructs were inspired
by the requirements of the irregular GPU algorithms proposed in the
literature, and that at least for these algorithms it is usually easy
to identify where to place the constructs.

@@

\paragraph{A changing number of workgroups}  Unlike in regular OpenCL,
the value returned by $\getnumgroups()$ is not fixed during the
lifetime of a cooperative kernel.  Although
$\getnumgroups()$ has atomic semantics, and so can be called at any
time to yield a well-defined value, the value returned by two
successive calls may differ if workgroups have executed $\offerkill$,
$\offerfork$ or $\resizingglobalbarrier$ in-between.  Because $\getglobalsize()$, which retrieves the total number of threads executing the kernel, is
a function of $\getnumgroups()$, the value returned by $\getglobalsize()$ is also subject to change.
The programmer
must therefore write their cooperative kernel in a manner that is
robust to changes in the values returned by these functions.

@@

For the work stealing example, there is no state to transmit since a
computation is entirely parameterised by a task, which is retrieved from
a task pool located in global memory. In the main loop, before trying to
obtain a task, we make the workgroup offer itself to be forked or killed
(lines~\ref{line:cwksteal:offerfork},\ref{line:cwksteal:offerkill}). Note
that a workgroup may be killed even if its associated task pool is not
empty, since remaining tasks could eventually be stealed by other
workgroups.  Care must be taken on computing pool identifiers: the
original version in \myfig\ref{fig:workstealing} can rely on the fact
that there are as many pools as workgroups, but here the number of
workgroups may change during execution. Therefore, the number of pools
must now be explicitely given as an argument
(line~\ref{line:cwksteal:argnumpool}) and used as an upper bound of pool
indexes when scanning pools in $\mathsf{more\_work}$
(line~\ref{line:cwksteal:morework}) and $\mathsf{pop\_or\_steal}$
(line~\ref{line:cwksteal:poporsteal}). Besides, computation of the
workgroup's pool index (line~\ref{line:cwksteal:getgroupid}) must not
only be bounded by the number of pools but also be moved inside the main
loop since a forked workgroup that starts to execute after $\offerfork$
must re-compute its pool index.

@@

The situation is more stable if a cooperative kernel does not directly
issue $\offerkill$ and $\offerfork$ calls directly.  In this case,
only calls to $\resizingglobalbarrier$ can cause the number of active
workgroups to change and, as discussed in \mysec\ref{sec:semantics},
this function causes synchronisation between all threads and causes
the number of active workgroups to change atomically.  At any point
during execution, the threads of a kernel are executing between some
pair of resizing barrier calls, which we call a \emph{resizing barrier
  intervals} (here we consider the kernel entry and exit points
conceptually to be special cases of resizing barriers).  The number of
workgroups executing the kernel is thus constant within each resizing
barrier interval, i.e.\ the result returned by repeated calls to
$\getnumgroups()$ will be identical during such an interval; the same holds for $\getglobalsize()$.

@@

Recall that the cooperative kernel would ideally utilise the $N$
workgroups requested upon kernel launch, in order to conduct its
processing efficiently.  The scheduler should thus aim to provide with
cooperative kernel with as many workgroups, up to this number, as
other constraints allow.  As a result, the scheduler should generally
accept an $\offerkill$ only if a compute unit is required for another
task, and should respond positively to $\offerfork$ calls by
scheduling new workgroups if compute units are available.  Our
implementation of cooperative scheduling (\mysec\ref{sec:implementation}) is as
generous as possible to the cooperative kernel in this regard.
However, these are not hard-and-fast rules: a scheduler might employ
heuristics based on historical workloads to avoid thrashing.  For
example, if graphics calls are known to arrive at a certain rate then
the scheduler might decide that it is worth accepting an $\offerkill$
from the cooperative kernel on the assumption that a graphics task
will be requested momentarily.  Similarly, the scheduler might provide
some, but not all, compute units to the cooperative kernel in response
to an $\offerfork$ call if it has reason to believe that the remaining
compute units are likely to be useful for other high priority tasks
that are likely to arrive imminently.

@@

\ADComment{Better section title?}
\HEComment{Dimensioning Non-persistent Workloads}

To assess how cooperative blocking algorithms can be, we fire lightweight
but still meaningful workloads that requires workgroups to be taken from
the blocking algorithms.

We tried to measure how much GPU computation is required to maintain
smooth animation on the screen during typical desktop usage. We model
these transient \HEComment{no sure about meaning if transient here}
using

The idea is
to run an infinite loop that launches a given kernel

@@
